---
title: "Introduction to reditR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to reditR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**reditR** identifies differentially RNA-edited (DRE) sites in bulk or
single-cell pseudobulk RNA-seq data. It takes per-site read-count tables
produced by upstream detection tools (e.g., SPRINT) and offers **three
independent statistical tests** -- a binomial GLMM, Fisher's exact test, and
a Wilcoxon rank-sum test -- each run on every site and independently
FDR-corrected. There is no gating of one test on another and no combined
verdict: you choose which test(s) suit your experimental design.

## End-to-end example

### 1. Read the data

```{r}
library(reditR)

ed <- read_editing_table(
  system.file("extdata", "example_editing.txt", package = "reditR")
)
mt <- read_metadata(
  system.file("extdata", "example_metadata.txt", package = "reditR"),
  reference_level = "control"
)

head(ed)
mt
```

### 2. Filter by coverage and cluster presence

`filter_editing_sites()` keeps rows where total ≥ `min_coverage` AND edited ≥
`min_edited`, then retains only sites present in ≥ `min_groups` samples.
The clustered subset (sites within `cluster_window` bp of a neighbour on the
same chromosome) is returned as `$clustered` — this is the standard input for
`differential_editing()`.

```{r}
# Write the example data to a temp file to demonstrate filter_editing_sites()
tmp_editing <- tempfile(fileext = ".txt")
write.table(as.data.frame(ed), tmp_editing, sep = "\t", row.names = FALSE, quote = FALSE)

filt <- filter_editing_sites(tmp_editing, min_coverage = 10, min_edited = 1, min_groups = 2)
cat("All filtered sites:", nrow(filt$all), "\n")
cat("Clustered sites:   ", nrow(filt$clustered), "\n")
```

### 3. Run differential analysis

`differential_editing()` accepts a file path to the filtered count table plus
a metadata file, so we write the example data to temp files for this demo.
The `test` argument selects which test(s) actually get computed -- here we
request all three so we can compare them side by side.

```{r}
tmp_meta <- tempfile(fileext = ".txt")
write.table(as.data.frame(mt), tmp_meta, sep = "\t", row.names = FALSE, quote = FALSE)

res <- differential_editing(
  data_path = tmp_editing,
  meta_path = tmp_meta,
  test      = c("glmm", "fisher", "wilcoxon"),
  verbose   = FALSE
)
print(res)
```

- **`chr1_100`** has a large control → diabetic editing shift (~6% vs ~23%) and
  is correctly flagged `GLMM_sig = TRUE`.
- **`chr1_200`** shows no difference and is `GLMM_sig = FALSE`.

Only asking for one test skips fitting the others entirely -- useful once
you've decided which test fits your design:

```{r}
res_wilcox_only <- differential_editing(
  data_path = tmp_editing,
  meta_path = tmp_meta,
  test      = "wilcoxon",
  verbose   = FALSE
)
names(res_wilcox_only)
```

### 4. Effect sizes

```{r}
eff <- editing_difference(tmp_editing, meta_path = tmp_meta)
print(eff)
```

## Choosing a test

reditR does not pick a test for you, because no single test was found to be
reliable across every design. The guidance below comes from running
permutation-based and parametric null-calibration simulations (relabelling
real replicates under the null, and simulating data under a known-null
generative model while sweeping between-replicate variance) across three
independent datasets:

- **GLMM** (`cbind(edited, unedited) ~ condition + (1|sample)`) is the
  primary test (Srivastava et al. 2017). It was severely anti-conservative
  -- up to ~9-10x the nominal false-positive rate -- in a design where
  between-replicate variance was high relative to the number of independent
  control replicates (e.g. a confounded or very-small-n design). In a
  balanced, adequately-replicated design the same test was only mildly
  anti-conservative.
- **Fisher's exact test** pools reads across samples per condition. It is
  independent of GLMM, but was anti-conservative to a similar degree
  whenever real between-replicate variance was present, regardless of
  dataset -- and unlike GLMM, adding more replicates did not fix it.
- **Wilcoxon rank-sum** on per-sample editing ratios was the most
  consistently well-calibrated test across every design checked, at the
  cost of lower power and a coarse p-value floor at small sample sizes
  (e.g. minimum achievable p-value ≈ 0.05 at n = 3 vs 3, ≈ 0.002 at n = 6
  vs 6).

Practical takeaway: a design with few, unbalanced, or confounded replicates
should lean on Wilcoxon; a well-powered balanced design can reasonably use
GLMM; Fisher should be treated cautiously whenever real between-replicate
variance is plausible. If you request more than one test, compare the
columns directly rather than trusting either test in isolation -- and build
your own combined criterion if you want one:

```{r}
res[res$GLMM_sig & res$Wilcox_sig, ]
```

## Simulation and validation

`validate_against_truth()` reports false-positive rate and power
independently for whichever test(s) were requested in `differential_editing()`
-- there is no combined metric, matching the independent-tests design above.

```{r}
sim <- simulate_editing_data(
  n_null   = 50,
  n_effects = c("0.10" = 20),
  seed     = 1
)
str(sim, max.level = 1)
```

```{r}
# Write sim data to temp files
tmp_sim_ed <- tempfile(fileext = ".txt")
tmp_sim_mt <- tempfile(fileext = ".txt")
write.table(as.data.frame(sim$editing),  tmp_sim_ed, sep = "\t", row.names = FALSE, quote = FALSE)
write.table(as.data.frame(sim$metadata), tmp_sim_mt, sep = "\t", row.names = FALSE, quote = FALSE)

res2 <- differential_editing(tmp_sim_ed, tmp_sim_mt,
                             reference_level = "control", case_level = "case",
                             test = c("glmm", "wilcoxon"), verbose = FALSE)
val  <- validate_against_truth(res2, sim$truth)
print(val)
```

```{r, include=FALSE}
# Clean up temp files
unlink(c(tmp_editing, tmp_meta, tmp_sim_ed, tmp_sim_mt))
```
